Spatiotemporal models of `r paste(params$species)` biomass with time-varying depth and trawl footprint

### 2020-02-19 analysis/SOPO/2-trawl-no-covs-model.Rmd
options to change knot number and update models for just rare speciesf

---
title: "Spatiotemporal models of `r paste(params$species)` biomass with time-varying depth and trawl footprint"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    # species: "Redbanded Rockfish"
    # species: "North Pacific Spiny Dogfish"
    # species: "Pacific Ocean Perch"
    # species: "Big Skate"
    # species: "Pacific Cod"
    #species: "Petrale Sole"
    species: "Starry Flounder"
    # species: "Bocaccio"
    # region: "All synoptic surveys"
    region: "Both odd year surveys"
    start_year: 2003
    knots: 250
    covariates: "" # +as.factor(ssid)
    #additional fixed-effect covariates
    covs: "-no-covs" # covariate string used to name model rds
    AR1: TRUE
    fixed_spatial: FALSE
    update_model: FALSE # TRUE
    update_model_check: FALSE #TRUE
    update_predictions: FALSE # TRUE
    update_index: TRUE
---

@@

```{r params}
species <- params$species
region <- params$region
covariates <- params$covariates
covs <- params$covs
knots <- params$knots
priors <- FALSE

paste("region =", region)
paste("model covariates =", covariates)
paste("model label =", covs)
paste("priors =", priors)
paste("knots =", knots)
```

@@

if(nrow(biomass)<2000) {stop("Need to recalculate split by maturity!")}

@@

rm(adult_biomass)
rm(imm_biomass)
rm(total_biomass)
```

Make mesh
```{r}
if (region == "Both odd year surveys") {
  spde <- sdmTMB::make_spde(data$X, data$Y, n_knots = 250)
}

@@

Run climate independent sdmTMB model
```{r }
if (params$update_model) {
  tictoc::tic()
  if (any(names(data) == "adult_density")) {
    adult_formula <- as.formula(paste(
      "adult_density ~ 0 + as.factor(year)", covariates, ""
    ))

@@

if(params$update_predictions) {
  # nd_all <- readRDS(paste0("data/nd_just_depth.rds"))
  # nd_all <- readRDS(paste0("data/nd_whole_coast_index.rds"))
  nd_all <- readRDS(paste0("data/nd_odd.rds"))

@@

### Check residuals
```{r}

if(params$update_model_check) {
try({

@@

```{r}
check_predictions <- readRDS(paste0(
  "data/", spp,
  "/check-mod-predictions-", spp, covs, "-", ssid_string, "-ar1-", params$AR1, "-reml.rds"
))

check_predictions <- filter(
  check_predictions,
  # ssid %in% c(1,3)) %>% filter (
  year > 2003
)

@@

g
```

@@

g2
```

```{r}
ggplot(check_predictions, aes(X, Y, colour = (residuals))) +
  geom_point() +
  scale_colour_gradient2() +
  scale_x_continuous(trans = "log10") +
  facet_wrap(~year)
```

@@

if (!exists("ad_predictions")) {
try({
  ad_predictions <- readRDS(paste0("data/", spp, 
    "/mod-mat-biomass-", spp, covs, "-", ssid_string, 
    "-ar1-", params$AR1, "-reml.rds"))
  })
  try({
    im_predictions <- readRDS(paste0("data/", spp, 
    "/mod-imm-biomass-", spp, covs, "-", ssid_string, 
    "-ar1-", params$AR1, "-reml.rds"))
  })
  if (!exists("ad_predictions")) {
    try({
      predicted <-  readRDS(paste0("data/", spp,
        "/model-total-biomass-", spp, covs, "-", ssid_string, 
        "-ar1-", params$AR1, "-reml.rds"))
    })
  }
}

@@

### Model summary
```{r}
 try({ ad_predictions$fit_obj})
 try({ im_predictions$fit_obj})
 try({ predicted$fitobj})
```

@@

```{r}
rm(ind)
rm(ind_imm)
if(params$update_index) {
if (exists("ad_predictions")) {
ind <- get_index(ad_predictions, bias_correct = FALSE)
ind$species <- species
ind$maturity <- "mature"
ind$ssid <- ssid_string
ind$covs <- covs

write.csv(ind, file = paste0("data/_indices/sopo-index-", 
  spp,"-", ssid_string, "-reml.csv"))

try ({
  ind_imm <- get_index(im_predictions, bias_correct = FALSE)
  ind_imm$species <- species
  ind_imm$maturity <- "immature"
  ind_imm$ssid <- ssid_string
  ind_imm$covs <- covs

  write.csv(ind_imm, file = paste0("data/_indices/sopo-index-imm-", 
    spp,"-", ssid_string, "-reml.csv"))
})
} else {
  ind <- get_index(predicted, bias_correct = FALSE)
  ind$species <- species
  ind$maturity <- "all"
  ind$ssid <- ssid_string
  ind$covs <- covs

  write.csv(ind, file = paste0("data/_indices/sopo-index-", 
    spp,"-", ssid_string, "-reml.csv"))
}
}

try({head(ind)})
try({head(ind_imm)})
```

@@

ind <- read.csv(paste0("data/_indices/sopo-index-", 
  spp,"-", ssid_string, "-reml.csv"))

ind %>% filter(year > 2004) %>%
ggplot(aes(year, est*scale)) + geom_line(col=ad_col) +
  geom_ribbon(aes(ymin = lwr*scale, ymax = upr*scale), fill=ad_col, alpha = 0.4) +
  xlab('Year') +  
  ylab('Mature biomass estimate (metric tonnes)') +
  gfplot::theme_pbs()
```

Add immature trend if available
```{r}
try({
  ind_imm <- read.csv(paste0("data/_indices/sopo-index-imm-", 
    spp,"-", ssid_string, "-reml.csv"))# %>% filter(year > 2004)
})

if (exists("ind_imm")) {
ratio <- mean(ind$est)/max(ind_imm$est)
imm_col <- "orangered"
ind %>% #filter(year > 2004) %>%
ggplot(aes(year, est*scale)) + geom_line(col = ad_col, ) +
  geom_ribbon(aes(ymin = lwr*scale, ymax = upr*scale), fill = ad_col, alpha = 0.4) +
# adding the relative humidity data, transformed to match roughly the range of the temperature
  geom_ribbon(aes(ymin = ind_imm$lwr*scale*ratio, ymax = ind_imm$upr*scale*ratio), 
    fill = "orangered", alpha = 0.4) + 
  geom_line(aes(ind_imm$year, ind_imm$est*scale*ratio), col = "orangered") +
# now adding the secondary axis, following the example in the help file ?scale_y_continuous and, reverting the above transformation
  scale_y_continuous(sec.axis = sec_axis(~./ratio, name = "Immature biomass estimate (metric tonnes)")) +
  xlab('Year') + 
  ylab('Mature biomass estimate (metric tonnes)')+
  gfplot::theme_pbs(base_size = 16)
}
```

>>>>

---
title: "Spatiotemporal models of `r paste(params$species)` biomass with time-varying depth and trawl footprint"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params:
    # species: "Redbanded Rockfish"
    # species: "North Pacific Spiny Dogfish"
    # species: "Pacific Ocean Perch"
    # species: "Pacific Sand Lance"
    # species: "Pacific Cod"
    # species: "Buffalo Sculpin"
    # species: "Cabezon"
    species: "Harlequin Rockfish"
    # region: "All synoptic surveys"
    region: "Both odd year surveys"
    start_year: 2003
    knots: 300
    covariates: "" # +as.factor(ssid)
    #additional fixed-effect covariates
    covs: "-no-covs-300" # covariate string used to name model rds
    AR1: TRUE
    fixed_spatial: FALSE
    update_model: FALSE #TRUE
    update_model_check: FALSE #TRUE
    update_predictions: FALSE #TRUE
    update_index: TRUE
---

@@

```{r params}
species <- params$species
region <- params$region
covariates <- params$covariates
covs <- params$covs
knots <- params$knots
priors <- FALSE
update_model <- params$update_model
update_model_check <- params$update_model_check
update_predictions <- params$update_predictions
update_index <- params$update_index

paste("region =", region)
paste("model covariates =", covariates)
paste("model label =", covs)
paste("priors =", priors)
```

@@

  positive_sets <- filter(biomass, density != 0)
  prop_pos <- round(nrow(positive_sets)/nrow(biomass), digits = 3)
      if (prop_pos < 0.1) {
        update_model <- T
        update_model_check <- T
        update_predictions <- T
        update_index <- T
        knots <- 150
        }
    if (prop_pos < 0.05) {
        update_model <- F
        update_model_check <- F
        update_predictions <- F
        update_index <- T
        knots <- 150
        }

# View(biomass)
if (nrow(biomass)<2000) {
  # update_model <- T
  # update_model_check <- T
  # update_predictions <- T
  # update_index <- T
  knots <- 100
  if(prop_pos < 0.05) { knots <- 50 }
  }
if(nrow(biomass)<1000) {stop("Need to recalculate split by maturity!")}

@@

rm(adult_biomass)
rm(imm_biomass)
rm(total_biomass)

```

Make mesh
```{r}
if (region == "Both odd year surveys") {
  spde <- sdmTMB::make_spde(data$X, data$Y, n_knots = knots)
}

@@

```{r}
paste("knots =", knots)
```

Run climate independent sdmTMB model
```{r }

if (update_model) {
  tictoc::tic()
  if (any(names(data) == "adult_density")) {
    adult_formula <- as.formula(paste(
      "adult_density ~ 0 + as.factor(year)", covariates, ""
    ))

@@

if(update_predictions) {

  # nd_all <- readRDS(paste0("data/nd_just_depth.rds"))
  # nd_all <- readRDS(paste0("data/nd_whole_coast_index.rds"))
  nd_all <- readRDS(paste0("data/nd_odd.rds"))

@@

### Check residuals
```{r}
if(update_model_check) {
try({

@@

```{r }

if(update_model_check) {
check_predictions <- readRDS(paste0(
  "data/", spp,
  "/check-mod-predictions-", spp, covs, "-", ssid_string, "-ar1-", params$AR1, "-reml.rds"
))

check_predictions <- filter(
  check_predictions,
  # ssid %in% c(1,3)) %>% filter (
  year > 2004
)

@@

g
}
```

@@

try ({g2})
```

```{r}
try({
  ggplot(check_predictions, aes(X, Y, colour = (residuals))) +
    geom_point() +
    scale_colour_gradient2() +
    scale_x_continuous(trans = "log10") +
    facet_wrap(~year)
})
```

Load predictions

@@

if (!exists("ad_predictions")) {
  try({
    ad_predictions <- readRDS(paste0("data/", spp, 
      "/mod-mat-biomass-", spp, covs, "-", ssid_string, 
      "-ar1-", params$AR1, "-reml.rds"))
  })

  if (!exists("ad_predictions")) {
     try({ ad_predictions <- readRDS(paste0("data/", spp,
      "/mod-mat-biomass-", spp, "-no-covs-", ssid_string,
      "-ar1-", params$AR1, "-reml.rds")) })
  }

  try({
    im_predictions <- readRDS(paste0("data/", spp, 
      "/mod-imm-biomass-", spp, covs, "-", ssid_string, 
      "-ar1-", params$AR1, "-reml.rds"))
  })
    if (!exists("im_predictions")) {
     try({ im_predictions <- readRDS(paste0("data/", spp,
      "/mod-imm-biomass-", spp, "-no-covs-", ssid_string,
      "-ar1-", params$AR1, "-reml.rds")) })
  }

  if (!exists("ad_predictions")) {
    try({
      predicted <-  readRDS(paste0("data/", spp,
        "/model-total-biomass-", spp, covs, "-", ssid_string, 
        "-ar1-", params$AR1, "-reml.rds"))
    })
    if (!exists("predicted")) {
     try({
       predicted <- readRDS(paste0("data/", spp,
      "/model-total-biomass-", spp, "-no-covs-", ssid_string,
      "-ar1-", params$AR1, "-reml.rds"))
       })
  }
  }
}

@@

### Model summary
```{r}
 try({ ad_predictions$fit_obj})
 try({ im_predictions$fit_obj})
 try({ predicted$fit_obj})
```

@@

```{r}
rm(ind)
rm(ind_imm)
library(data.table)
if(update_index) {

if (exists("ad_predictions")) {

  ind <- get_index(ad_predictions, bias_correct = FALSE)
  ind$species <- species
  ind$maturity <- "mature"
  ind$ssid <- ssid_string
  ind$covs <- covs
  ind <- as.data.frame(ind)
  write.csv(ind, file = paste0("data/_indices/sopo-index-", 
    spp,"-", ssid_string, "-", covs, "-reml.csv"), row.names = FALSE)

  try ({
    ind_imm <- get_index(im_predictions, bias_correct = FALSE)
    ind_imm$species <- species
    ind_imm$maturity <- "immature"
    ind_imm$ssid <- ssid_string
    ind_imm$covs <- covs
    ind_imm <- as.data.frame(ind_imm)
    write.csv(ind_imm, file = paste0("data/_indices/sopo-index-imm-", 
      spp,"-", ssid_string, "-", covs, "-reml.csv"), row.names = FALSE)
  })

} else {
  ind <- get_index(predicted, bias_correct = FALSE)
  ind$species <- species
  ind$maturity <- "all"
  ind$ssid <- ssid_string
  ind$covs <- covs

  ind <- as.data.frame(ind)

  write.csv(ind, file = paste0("data/_indices/sopo-index-", 
    spp,"-", ssid_string, "-", covs, "-reml.csv"), row.names = FALSE)
}
try({ glimpse(ind_imm) })
try({ glimpse(ind) })
}  
```

@@

ind <- read.csv(paste0("data/_indices/sopo-index-", 
  spp,"-", ssid_string, "-", covs, "-reml.csv"))

ind %>% filter(year > 2004) %>%
ggplot(aes(year, est*scale)) + geom_line(col=ad_col) +
  geom_ribbon(aes(ymin = lwr*scale, ymax = upr*scale), fill=ad_col, alpha = 0.4) +
  xlab('Year') +  
  ylab('Mature biomass estimate (metric tonnes)') +
  gfplot::theme_pbs() +
    ggtitle(paste0("", species, " "))
```

Add immature trend if available
```{r}
try({
  ind_imm <- read.csv(paste0("data/_indices/sopo-index-imm-", 
    spp,"-", ssid_string, "-", covs, "-reml.csv")) %>% filter(year > 2004)
})

if (exists("ind_imm")) {
ratio <- mean(ind$est)/max(ind_imm$est)
imm_col <- "orangered"

ind %>% filter(year > 2004) %>%
ggplot(aes(year, est*scale)) + geom_line(col = ad_col, ) +
  geom_ribbon(aes(ymin = lwr*scale, ymax = upr*scale), fill = ad_col, alpha = 0.4) +
# adding the relative humidity data, transformed to match roughly the range of the temperature
  geom_ribbon(aes(ymin = ind_imm$lwr*scale*ratio, ymax = ind_imm$upr*scale*ratio), 
    fill = "orangered", alpha = 0.4) + 
  geom_line(aes(ind_imm$year, ind_imm$est*scale*ratio), col = "orangered") +
# now adding the secondary axis, following the example in the help file ?scale_y_continuous and, reverting the above transformation
  scale_y_continuous(sec.axis = sec_axis(~./ratio, name = "Immature biomass estimate (metric tonnes)")) +
  xlab('Year') + 
  ylab('Mature biomass estimate (metric tonnes)')+
  gfplot::theme_pbs(base_size = 16)
}
```
